---
title: "K08: DGE by EdgeR-pseudobulk based on Pf status with p53 CAMERA"
author: "Srija"
date: "2023-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load packages
```{r}
library(Seurat) 
library(patchwork)
library(SeuratDisk)
library(tidyverse)
library(magrittr)
library(googledrive)
library(ggpubr)
library(scales)
library(RColorBrewer)
library(viridis)
library(scRNAseq)
library(scran)
library(Glimma)
library(edgeR)
library(SingleCellExperiment)
```

```{r}
# Authenticate Google Drive (will ask you to log in)
drive_auth()

# Use the file ID to get the file
file_id <- as_id("1swCwEuB3ez48nDHl1jXPTBGdhMPIzX0m")

# Download to a temp path
temp_rds <- tempfile(fileext = ".rds")
drive_download(file = file_id, path = temp_rds, overwrite = TRUE)

```

```{r}
refmapped.all <- readRDS("refmapped_all.rds")
```

# Let's look at the cells
```{r}
dimplot_hires<-DimPlot(refmapped.all, reduction = "ref.umap", group.by = "predicted.celltype.l2", pt.size =0.35, label = T, raster = F) +
    NoLegend() +
     ggtitle("K08_CITEseq: High resolution celltype annotation")

# Very few monocytes (n=133, we will have to adjust cutoffs downstream)
tab<-table(refmapped.all$Sample.ID, refmapped.all$predicted.celltype.l2)
colSums(tab)
```

### subset and extract the cells from a Seurat object

```{r, subset and extract from seurat}
# Extract raw counts and metadata to create SingleCellExperiment object
counts <- refmapped.all@assays$RNA@counts 
metadata <- refmapped.all@meta.data
# Set up metadata as desired for aggregation and DE analysis
refmapped.all <- SetIdent(refmapped.all, value = factor(gsub("\\_"," ", refmapped.all@meta.data$predicted.celltype.l2)))
metadata$cluster_id <- factor(refmapped.all@active.ident)
metadata <- metadata %>%
  dplyr::rename(sample_id = "Sample.ID",
                class = "pfstatus") %>%
  mutate(class = factor(tolower(class))) %>%
  mutate(experiment_mode = factor(gsub(".*a\\_", "", sample_id))) %>%
  mutate(sample_id = factor(gsub("\\_.*", "", tolower(sample_id))))
# Create single cell experiment object
sce <- SingleCellExperiment(assays = list(counts = counts), 
                           colData = metadata)
#change colnames of coldata to more r friendy names
colnames(colData(sce)) <- gsub("\\.","_",  colnames(colData(sce)))
# Identify groups for aggregation of counts
groups <- colData(sce)[, c("cluster_id", "sample_id", "predicted_celltype_l2")] 
```


### Identify highly variable genes for creating MDS plot using OSCA procedure

```{r follow OSCA procedure}
#uses scuttle, scran
sce <- scuttle::logNormCounts(sce)
var_mod <- scran::modelGeneVar(sce)
hvg_genes <- scran::getTopHVGs(var_mod, n=500)
hvg_sce <- sce[hvg_genes, ]
hvg_sce <- logNormCounts(hvg_sce)
```


### acquiring necessary metrics for aggregation across cells in a sample

```{r, acquire metrics}
# Named vector of cluster names
kids <- purrr::set_names(levels(sce$cluster_id))
kids
# Total number of clusters
nk <- length(kids)
nk
# Named vector of sample names
sids <- purrr::set_names(levels(sce$sample_id))
# Total number of samples 
ns <- length(sids)
ns
# Generate sample level metadata
## Determine the number of cells per sample
table(sce$sample_id)
table(sce$experiment_mode)
## Turn named vector into a numeric vector of number of cells per sample
n_cells <- as.numeric(table(sce$sample_id))
## Determine how to reorder the samples (rows) of the metadata to match the order of sample names in sids vector
m <- match(sids, sce$sample_id)
## Create the sample level metadata by combining the reordered metadata with the number of cells corresponding to each sample.
ei <- data.frame(colData(sce)[m, ], 
                  n_cells, row.names = NULL) %>% 
                dplyr::select(-"cluster_id")
ei %>%
  dplyr::select(class, sample_id, n_cells)
```



### Remove low quality genes 

(re-read vignette to important points)

```{r, remove low quality using scater}
#skip because can't install scater
# Perform QC if not already performed
dim(sce)
# Calculate quality control (QC) metrics
#sce <- calculateQCMetrics(sce)
# Get cells w/ few/many detected genes
#sce$is_outlier <- isOutlier(
#        metric = sce$total_features_by_counts,
#        nmads = 2, type = "both", log = TRUE)
# Remove outlier cells
#sce <- sce[, !sce$is_outlier]
```


```{r, remove lowly expressed genes}
## Remove lowly expressed genes which have less than 20 cells with any counts
dim(sce)
sce <- sce[rowSums(counts(sce) > 1) >= 20, ]
dim(sce)
```

## Count aggregation to sample level

```{r, count aggregation to sample level}
# Aggregate the counts per sample_id and cluster_id
# Subset metadata to only include the cluster and sample IDs to aggregate across
groups <- colData(sce)[, c("cluster_id", "sample_id")]
# Aggregate across cluster-sample groups
pb <- Matrix.utils::aggregate.Matrix(t(counts(sce)), 
                       groupings = groups, fun = "sum") 
class(pb)
dim(pb)
pb[1:6, 1:6]
```

### Split data by cell type

```{r split data by cell type}
# Not every cluster is present in all samples; create a vector that represents how to split samples
splitf <- sapply(stringr::str_split(rownames(pb), 
                                    pattern = "_(?!.*_)",
                                    n = 2), 
                 `[`, 1)
# Turn into a list and split the list into components for each cluster and transform, so rows are genes and columns are samples and make rownames as the sample IDs
pb <- split.data.frame(pb, 
                       factor(splitf)) %>%
        lapply(function(u) 
                set_colnames(t(u), 
                             stringr::str_extract(rownames(u), "(?<=_)[:alnum:]+")))
class(pb)
# Explore the different components of list
str(pb)
```

### Check counts per sample for each cluster

```{r, check counts per sample for each cluster}
# Print out the table of cells in each cluster-sample group
options(width = 100)
table(sce$cluster_id, sce$sample_id)
colSums(table(sce$cluster_id, sce$sample_id))
colSums(t(table(sce$cluster_id, sce$sample_id)))
cell_types_above_100 <- levels(sce$cluster_id)[colSums(t(table(sce$cluster_id, sce$sample_id))) > 100] #limit to cell types with enough cells
```

## Prepare sample-level metadata for DE anaylsis

```{r, prepare sample level metadata}
# Get sample names for each of the cell type clusters
# prep. data.frame for plotting
get_sample_ids <- function(x){
        pb[[x]] %>% 
                colnames()
}
de_samples <- map(1:length(kids), get_sample_ids) %>%
        unlist()
# Get cluster IDs for each of the samples
samples_list <- map(1:length(kids), get_sample_ids)
get_cluster_ids <- function(x){
        rep(names(pb)[x], 
            each = length(samples_list[[x]]))
}
de_cluster_ids <- map(1:length(kids), get_cluster_ids) %>%
        unlist()
# Create a data frame with the sample IDs, cluster IDs and condition
gg_df <- data.frame(cluster_id = de_cluster_ids,
                    sample_id = de_samples)
gg_df <- left_join(gg_df, ei[, c("sample_id", "class")]) 
metadata <- gg_df %>%
        dplyr::select(cluster_id, sample_id, class) 
        
metadata 
```

### Subsetting dataset to cluster(s) of interest

```{r subset to clusters of interest}
# Generate vector of cluster IDs
clusters <- levels(factor(metadata$cluster_id))
clusters
```

Subset on celltype of interest and run edgeR as a loop
```{r}

```

https://bioconductor.org/packages/release/bioc/vignettes/Glimma/inst/doc/single_cell_edger.html

```{r subset on celltype of interest then run edgeR all in a loop}
p53_gene_list <- readRDS("p53_geneset.rds")

# Define input variables (assume already computed)
# cell_types_above_100 = names of cell types with ≥100 cells
# pb = pseudobulk counts per cell type
# metadata = sample metadata containing class (e.g., pf_plus/pf_neg)

# Initialize outputs
pb_mroast_p53 <- pb_camera_p53 <- toptags_tbl <- list()
pb_qlf <- pb_qlfit <- pb_lrt <- pb_fit <- contr <- design <- dge_pb <- counts <- cluster_counts <- cluster_metadata <- 
  sapply(cell_types_above_100, function(x) NULL)

for(i in cell_types_above_100){
  cluster_metadata[[i]] <- metadata[which(metadata$cluster_id == i), ]
  rownames(cluster_metadata[[i]]) <- cluster_metadata[[i]]$sample_id
  counts[[i]] <- pb[[i]]
  cluster_counts[[i]] <- data.frame(counts[[i]][, colnames(counts[[i]]) %in% rownames(cluster_metadata[[i]])])

  if(all(rownames(cluster_metadata[[i]]) == colnames(cluster_counts[[i]]))){
    print(paste0("Sample/Count match: ", i))
  }

  dge_pb[[i]] <- DGEList(counts = cluster_counts[[i]], samples = cluster_metadata[[i]], group = cluster_metadata[[i]]$class)
  dge_pb[[i]] <- calcNormFactors(dge_pb[[i]])
  design[[i]] <- model.matrix(~0 + class, data = dge_pb[[i]]$samples)
  colnames(design[[i]]) <- make.names(gsub("class", "", colnames(design[[i]])))
  dge_pb[[i]] <- estimateDisp(dge_pb[[i]], design[[i]])
  contr[[i]] <- makeContrasts(pf_plus - pf_neg, levels = design[[i]])

  pb_fit[[i]] <- glmFit(dge_pb[[i]], design[[i]], robust = TRUE)
  pb_lrt[[i]] <- glmLRT(pb_fit[[i]], contrast = contr[[i]])
  pb_qlfit[[i]] <- glmQLFit(dge_pb[[i]], design[[i]], robust = TRUE)
  pb_qlf[[i]] <- glmQLFTest(pb_qlfit[[i]], contrast = contr[[i]])

  # Run mroast and camera using your p53 gene list
  pb_mroast_p53[[i]] <- mroast(dge_pb[[i]], index = gene_list, design = design[[i]], contrast = contr[[i]])
  pb_camera_p53[[i]] <- camera(dge_pb[[i]], index = gene_list, design = design[[i]], contrast = contr[[i]])

  # Top DE genes
  toptags_tbl[[i]] <- topTags(pb_qlf[[i]], n = nrow(pb_qlf[[i]]))
}


```



#Plot CAMERA and mroast results
```{r}
# Required libraries
library(tibble)
library(purrr)

# Plot settings
basetextsize <- 12
myfont <- "sans"
bubble_max_size <- 24
mycolors <- c("Up" = "#F21A00", "Down" = "#3B9AB2")

# Combine CAMERA results
camera_bound <- imap(pb_camera_p53, ~ .x %>%
                       rownames_to_column("pathway")) %>%
  bind_rows(.id = "cell_type")

# Filter significant pathways
camera_filtered <- camera_bound %>%
  mutate(neglog10fdr = -log10(FDR)) %>%
  arrange(desc(neglog10fdr)) %>%
  filter(FDR < 0.2) %>%
  filter(cell_type != "dnT")  # Optional

# Create bubble plot
plot_camera_bubble <- ggplot(camera_filtered, aes(x = cell_type, y = pathway, color = Direction)) +
  geom_point(aes(size = neglog10fdr), alpha = 0.65) +
  scale_color_manual(values = mycolors, name = "Regulation") +
  scale_size_area(name = expression(-log[10]~FDR), max_size = bubble_max_size) +
  theme_minimal(base_family = myfont, base_size = basetextsize) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom",
    strip.background = element_blank()
  ) +
  coord_flip()

# Print plot
print(plot_camera_bubble)

# Save plot to file (optional)
ggsave("camera_p53_bubbleplot.pdf", plot_camera_bubble, width = 12, height = 7)


```

## Fischer p53 targets (Plot only mroast here)
```{r}
# Plot settings
basetextsize <- 12
myfont <- "sans"
bubble_max_size <- 24
mycolors <- c("Up" = "#F21A00", "Down" = "#3B9AB2")

# Combine mroast results
roast_bound <- imap(pb_mroast_p53, ~ .x %>%
                      rownames_to_column("pathway")) %>%
  bind_rows(.id = "cell_type")

# Filter significant results
roast_filtered <- roast_bound %>%
  mutate(neglog10fdr = -log10(FDR)) %>%
  arrange(desc(neglog10fdr)) %>%
  filter(FDR < 0.2) %>%             # Adjust FDR cutoff as needed
  filter(cell_type != "dnT")         # Optional: remove dnT if needed

# Create bubble plot
plot_mroast_p53 <- ggplot(roast_filtered, aes(x = cell_type, y = pathway, color = Direction)) +
  geom_point(aes(size = neglog10fdr), alpha = 0.65) +
  scale_color_manual(values = mycolors, name = "Regulation") +
  scale_size_area(name = expression(-log[10]~FDR), max_size = bubble_max_size) +
  theme_minimal(base_family = myfont, base_size = basetextsize) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom",
    strip.background = element_blank()
  ) +
  coord_flip()

# Show plot
print(plot_mroast_p53)

# Optional: Save as PDF
ggsave("mroast_p53_bubbleplot.pdf", plot_mroast_p53, width = 14, height = 14)

```







```{r}

# UMAP Plot - colored by cell type
umap_camera_plot <- DimPlot(
  object = refmapped.all,
  reduction = "ref.umap",                  # Use your existing ref.umap
  group.by = "predicted.celltype.l2",       # Cell type column
  label = TRUE,                             # Add cell type labels
  repel = TRUE,                             # Avoid overlap of labels
  pt.size = 0.35                            # Size of points
) +
  ggtitle("UMAP of PBMC Cell Types") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"                # Hide the legend since you have labels
  )

# Print UMAP
print(umap_camera_plot)

# Save UMAP as a PDF (optional)
ggsave("camera_umap_plot.pdf", umap_camera_plot, width = 7, height = 5)

```



```{r}
# Create a bright color palette using viridis
# Option "D" gives bright colors; direction = -1 makes it more colorful
bright_colors <- viridis(
  n = length(unique(refmapped.all$predicted.celltype.l2)), 
  option = "D", 
  direction = -1
)

# UMAP Plot - using bright viridis colors
umap_camera_plot <- DimPlot(
  object = refmapped.all,
  reduction = "ref.umap",
  group.by = "predicted.celltype.l2",
  label = TRUE,
  repel = TRUE,
  pt.size = 0.35
) +
  scale_color_manual(values = bright_colors) +   # Apply the viridis color palette
  ggtitle("UMAP of PBMC Cell Types") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"
  )

# Print UMAP
print(umap_camera_plot)

# Save UMAP as a PDF (optional)
ggsave("camera_umap_plot_viridis.pdf", umap_camera_plot, width = 7, height = 5)

```

```{r}
# Create bright viridis colors
celltypes <- unique(refmapped.all$predicted.celltype.l2)
bright_colors <- viridis(
  n = length(celltypes), 
  option = "D", 
  direction = -1
)
names(bright_colors) <- celltypes  # Name the colors for proper matching

# UMAP plot with cell type names as legend (not labels on the plot)
umap_camera_plot <- DimPlot(
  object = refmapped.all,
  reduction = "ref.umap",
  group.by = "predicted.celltype.l2",
  label = FALSE,       # No labels on points
  pt.size = 0.35
) +
  scale_color_manual(values = bright_colors, name = "Cell Types") +  # Legend on side
  ggtitle("UMAP of PBMC Cell Types") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right",         # Show cell names as a legend on the side
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold")
  )

# Show the plot
print(umap_camera_plot)

# Save to file (optional)
ggsave("camera_umap_plot_with_legend.pdf", umap_camera_plot, width = 8, height = 6)

```

